
---
title: 'Data 624 - Homework #8'
author: "Wei Zhou"
date: '2020-11-15'
---


Exercises 7.2 & 7.5  from the K&J book. 

```{r load.requirements, warning = FALSE, message = FALSE}
#clear the workspace
rm(list = ls())
#load req's packages
library(mlbench)
library(caret)
library(knitr)
library(AppliedPredictiveModeling)
library(corrplot)
```

## 7.2

Friedman (1991) introduced several benchmark data sets create by simulation.
One of these simulations used the following nonlinear equation to
create data:

$y=10sin(\pi x_1x_2)+20(x_3-0.5)^2 + 10x_4+5x_5+N(0,\sigma^2)$

where the x values are random variables uniformly distributed between [0, 1]
(there are also 5 other non-informative variables also created in the simulation).
The package mlbench contains a function called mlbench.friedman1 that
simulates these data:

```{r}
set.seed(200)
trainingData <- mlbench.friedman1(200, sd = 1)
## We convert the 'x' data from a matrix to a data frame
## One reason is that this will give the columns names.
trainingData$x <- data.frame(trainingData$x)
## Look at the data using
featurePlot(trainingData$x, trainingData$y)
## or other methods.
## This creates a list with a vector 'y' and a matrix
## of predictors 'x'. Also simulate a large test set to
## estimate the true error rate with good precision:
testData <- mlbench.friedman1(5000, sd = 1)
testData$x <- data.frame(testData$x) 
```
Tune several models on these data. For example:

```{r}
knnModel <- train(x = trainingData$x, 
                  y = trainingData$y,
                  method = "knn",
                  preProcess = c("center", "scale"),
                  tuneLength = 10)
knnModel
```



```{r}
knnPred <- predict(knnModel, newdata = testData$x)
postResample(pred = knnPred, obs = testData$y)
```

Which models appear to give the best performance? Does MARS select the
informative predictors (those named $X1-X5$)?

### MARS:

```{r}
set.seed(19)  #set the see for the 7.2
grid.MARS <- expand.grid(degree =1:2, nprune=seq(2,14,by=2))
model.MARS <- train(x = trainingData$x, 
                    y = trainingData$y, 
                    method='bagEarth', 
                    tuneGrid = grid.MARS, 
                    trControl = trainControl(method = "cv"))
model.MARS
pred.MARS <- predict(model.MARS, newdata = testData$x)
postResample(pred = pred.MARS, obs = testData$y)
```

### Neural Network:

```{r}
grid.NN <- expand.grid(size = seq(from = 1, to = 5, by = 1),
                        decay = seq(from = 0.01, to = 0.1, by = 0.01), 
                        bag = F)
maxW.NN <- 5 * (ncol(trainingData$x) + 1) + 5 + 1
model.NN <- train(x = trainingData$x,
                    y = trainingData$y,
                    method = "avNNet",
                    preProcess = c("center", "scale"),
                    tuneGrid = grid.NN,
                    trControl = trainControl(method = "cv"),
                    linout = TRUE,
                    trace = FALSE,
                    MaxNWts = maxW.NN,
                    maxit = 500)
model.NN
pred.NN <- predict(model.NN, newdata = testData$x)
postResample(pred = pred.NN, obs = testData$y)
```


### SVM:

```{r}
model.SVM <- train(x = trainingData$x,
                   y = trainingData$y,
                   method='svmRadial',
                   tuneLength = 9,
                   trControl = trainControl(method = "cv"))
model.SVM
pred.SVM <- predict(model.SVM, newdata = testData$x)
postResample(pred = pred.SVM , obs = testData$y)
```

### Summary

First we will look at model performance:

```{r}
model.perf <- cbind(data.frame(postResample(pred = knnPred, obs = testData$y)),
                    data.frame(postResample(pred = pred.MARS, obs = testData$y)),
                    data.frame(postResample(pred = pred.NN, obs = testData$y)),
                    data.frame(postResample(pred = pred.SVM, obs = testData$y)))
colnames(model.perf) <- c("KNN","MARS","NN","SVM")
kable(round(model.perf,2))
```

For this particular problem, we can see that the MARS model the superior performer.  MARS shows the smallest error stats and the higest Rsquared. 


```{r}
var.importance <- cbind(varImp(knnModel)[[1]],
                        varImp(model.NN)[[1]],
                        varImp(model.SVM)[[1]])
colnames(var.importance) <- c("KNN","NN","SVM")
kable(round(var.importance))
```

All of the models correctly identify $X1-X5$ as the important variables while almost identifying $X6-X10$ as not important.

Interestingly, MARS performs the best and also shows 2 distinct differences:  For the MARS model,   $X3$ is significantly over-weight as compared to other models and $X4$ is signifcantly underweight vs. the other models. 

 
## 7.5

Exercise 6.3 describes data for a chemical manufacturing process. Use
the same data imputation, data splitting, and pre-processing steps as before
and train several nonlinear regression models.

```{r}
set.seed(19) #reset the seed for Q 7.5 
data(ChemicalManufacturingProcess)
chem <- ChemicalManufacturingProcess
#impute using knn
chem.imp <- preProcess(chem[,2:ncol(chem)], method=c('knnImpute'))
chem <- cbind(chem$Yield,predict(chem.imp, chem[,2:ncol(chem)]))
colnames(chem)[1] <- "Yield"
#split 70/30
n <-  floor(0.70 * nrow(chem))
idx <- sample(seq_len(nrow(chem)), size = n)
train <- chem[idx, ]
test <- chem[-idx, ]
```


```{r, warning=FALSE}
knnModel <- train(x=train[,-1],
                  y=train$Yield,
                  method = "knn",
                  preProcess = c("center", "scale"),
                  tuneLength = 10)
knnModel
knnPred <- predict(knnModel, newdata = test[,-1])
postResample(pred = knnPred, obs = test$Yield)
grid.MARS <- expand.grid(degree =1:2, nprune=seq(2,14,by=2))
model.MARS <- train(x=train[,-1],
                    y=train$Yield, 
                    method='bagEarth', 
                    tuneGrid = grid.MARS, 
                    trControl = trainControl(method = "cv"))
model.MARS
pred.MARS <- predict(model.MARS, newdata = test[,-1])
postResample(pred = pred.MARS, obs = test$Yield)
grid.NN <- expand.grid(size = seq(from = 1, to = 5, by = 1),
                        decay = seq(from = 0.01, to = 0.1, by = 0.01), 
                        bag = F)
maxW.NN <- 5 * (ncol(train[,-1]) + 1) + 5 + 1
model.NN <- train(x=train[,-1],
                 y=train$Yield,
                    method = "avNNet",
                    preProcess = c("center", "scale"),
                    tuneGrid = grid.NN,
                    trControl = trainControl(method = "cv"),
                    linout = TRUE,
                    trace = FALSE,
                    MaxNWts = maxW.NN ,
                    maxit = 500)
model.NN
pred.NN <- predict(model.NN, newdata = test[,-1])
postResample(pred = pred.NN, obs = test$Yield)
model.SVM <- train(x=train[,-1],
                   y=train$Yield,
                   method='svmRadial',
                   tuneLength = 9,
                   trControl = trainControl(method = "cv"))
model.SVM
pred.SVM <- predict(model.SVM, newdata = test[,-1])
postResample(pred = pred.SVM , obs = test$Yield)
```

### A 
Which nonlinear regression model gives the optimal resampling and test set performance?

```{r}
model.perf <- cbind(data.frame(postResample(pred = knnPred, obs =test$Yield)),
                    data.frame(postResample(pred = pred.MARS, obs = test$Yield)),
                    data.frame(postResample(pred = pred.NN, obs = test$Yield)),
                    data.frame(postResample(pred = pred.SVM, obs = test$Yield)))
colnames(model.perf) <- c("KNN","MARS","NN","SVM")
kable(round(model.perf,2))
```

### B 
Which predictors are most important in the optimal nonlinear regression
model? Do either the biological or process variables dominate the
list? How do the top ten important predictors compare to the top ten
predictors from the optimal linear model?

```{r}
var.importance <- cbind(varImp(knnModel)[[1]],
                        varImp(model.NN)[[1]],
                        varImp(model.SVM)[[1]])
colnames(var.importance) <- c("KNN","NN","SVM")
kable(round(var.importance))