---
title: "Data624 HW1"
author: "Wei Zhou"
date: "9/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 2.10 Exercise
```{r}
library(tidyr)
library(knitr)
library(utils)
library(ggplot2)
library(readxl)
library(forecast)
library(fpp2)
```

### 2.1 Use the help function to explore what the series gold, woolyrnq and gas represent.

a. Use autoplot() to plot each of these in separate plots.

```{r}
autoplot(gold) +
  ggtitle("Daily morning gold prices") +
  xlab("1 January 1985 to 31 March 1989") +
  ylab("US Dollars")

autoplot(woolyrnq) +
  ggtitle("Quarterly production of woollen yarn in Australia") +
  xlab("Mar 1965 to Sep 1994") +
  ylab("Tonnes")

autoplot(gas) +
  ggtitle("Australian monthly gas production") +
  xlab("1956 to 1995") +
  ylab("production")

```

b. What is the frequency of each series? Hint: apply the frequency() function.

```{r}
frequency(gold)
frequency(gas)
frequency(woolyrnq)

```

c. Use which.max() to spot the outlier in the gold series. Which observation was it?

```{r}

which.max(gold)
which.max(gas)
which.max(woolyrnq)
```

### 2.2 Download the file tute1.csv from the book website, open it in Excel (or some other spreadsheet application), and review its contents. You should find four columns of information. Columns B through D each contain a quarterly series, labelled Sales, AdBudget and GDP. Sales contains the quarterly sales for a small company over the period 1981-2005. AdBudget is the advertising budget and GDP is the gross domestic product. All series have been adjusted for inflation.

```{r}
file1 = "https://raw.githubusercontent.com/vsinha-cuny/data624/master/hw1/tute1.csv"
tute1 = read.csv(file = file1, header = T)

mytimeseries <- ts(tute1[,-1], start=1981, frequency=4)
autoplot(mytimeseries, facets=TRUE)
```
```{r}
autoplot(mytimeseries, facets=FALSE)

```
### 2.3 Download some monthly Australian retail data from the book website. These represent retail sales in various categories for different Australian states, and are stored in a MS-Excel file. Can you spot any seasonality, cyclicity and trend? What do you learn about the series?

```{r}
library(openxlsx)
file2 = "https://raw.githubusercontent.com/vsinha-cuny/data624/master/hw1/retail.xlsx"
retaildata <- read.xlsx(file2, sheet=1, startRow=2)
myts <- ts(retaildata[,"A3349873A"], frequency=12, start=c(1982,4))
autoplot(myts) +
    ggtitle("Australian retail data")
```
Seasonality: There is a large jump in sales starting in October and lasting till end of December.

```{r}
ggseasonplot(myts)

```
```{r}
ggsubseriesplot(myts)
```

The subseries plot shows the mean values of the time series. In this case we see that the values are in an upward trend from October to December.

```{r}
gglagplot(myts, lags=12)

```

From the lagplot we see that there is a strong correlation for all lag values. It is strongest for lag=12.

```{r}
ggAcf(myts)


```
The ACF plot shows strong autocorrelation. The seasonality is reflected in the strongest ACF being observed at lag=12.

### 2.6 Use the following graphics functions: autoplot(), ggseasonplot(), ggsubseriesplot(), gglagplot(), ggAcf() and explore features from the following time series: hsales, usdeaths, bricksq, sunspotarea, gasoline.

```{r}
autoplot(hsales) +
    ggtitle("Monthly sales of new one-family houses sold in the USA since 1973")
```

```{r}
ggseasonplot(hsales)

```
```{r}
ggsubseriesplot(hsales)

```
```{r}
gglagplot(hsales, lags=12)

```

```{r}
ggAcf(hsales)

```
```{r}
autoplot(usdeaths) +
    ggtitle("Monthly accidental deaths in USA")

ggseasonplot(usdeaths)

ggsubseriesplot(usdeaths)

gglagplot(usdeaths, lags=12)

ggAcf(usdeaths)

```


```{r}
autoplot(bricksq) +
    ggtitle("Australian quarterly clay brick production: 1956–1994")

ggseasonplot(bricksq)

ggsubseriesplot(bricksq)

gglagplot(bricksq, lags=12)

ggAcf(bricksq)

```

```{r}
autoplot(sunspotarea) +
    ggtitle("Annual average sunspot area (1875-2015)")
    
# ggseasonplot(sunspotarea)
# 
# ggsubseriesplot(sunspotarea)
# 
# gglagplot(sunspotarea, lags=12)
# 
# ggAcf(sunspotarea)

```

```{r}
autoplot(gasoline) +
    ggtitle("US finished motor gasoline product supplied")
    
ggseasonplot(gasoline)

# ggsubseriesplot(gasoline)

gglagplot(gasoline, lags=12)

ggAcf(gasoline)

```

